---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# TAPACLOTH

<!-- badges: start -->
[![R-CMD-check](https://github.com/caravagnalab/TAPACLOTH/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/caravagnalab/TAPACLOTH/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

TAPACLOTH is a tool that classifies somatic mutation calls from
    targeted-panel sequencing as either subclonal, clonal heterozygous or clonal 
    LOH, thorugh a statistical hypothesis test based on a Binomial or Beta-Binomial 
    modeling of the sequencing read count process. Additionally, 
    TAPACLOTH can be used to infer sample purity. 
    Since the two processes are independent, one can first estimate sample purity
    and then use it as input for the classification task.

## Installation

You can install the development version of TAPACLOTH from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("caravagnalab/TAPACLOTH")
```

## Example

This is a basic example of running the classification:

```{r}
library(TAPACLOTH)
## basic example code
x = dplyr::tibble(sample = "test",
           gene = "test gene",
           nv = 50,
           dp = 100,
           VAF = 0.5,
           purity = 1
)

run_classifier(
  data = x,
  alpha_level = 1e-3,
  model = "BetaBinomial",
  rho = 0.01
)
```
