---
title: "Classify karyotype and multiplicity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Classify clonality and zygosity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options(crayon.enabled=TRUE)
```

```{r setup}
library(dplyr)
library(TAPACLOTH)
```

We work with the [example TAPACLOTH `data`](TAPACLOTH.html).

```{r, include=FALSE}
print(example_data)
```

# Exact read-counts based test

TAPACLOTH provides an exact test (with a p-value) to assess karyotype and multipilcity from read-count data.

This test is based on a model of the sequencing outcome following either a Binomial or a Beta-Binomial process, in which the number of reads with a variant represents the number of successes over a total number of trials corresponding to coverage. 

A clonal variant affecting a genome site of ploidy $p$, with multiplicity $m \in \{1,2\}$ in a sample of purity $\pi$
has an expected VAF of
\[
\mathbb{E}[\text{VAF}] = \frac{m\pi}{p\pi + 2left(1-\pi\right)}
\] 

In the (Beta-)Binomial process this represents the success probability of the counting process. Therefore, if we assume as null model each possible combination of karyotype and multiplicity, and consider for every input mutation

- $NV$ reads with the variant at the locus;
- $DP$ coverage at the locus;
- sample purity $\pi$;

then the expected number of reads with the variant follows
\[
P(X = NV) = Bin\left(\cdot \;\large\mid\;\normalsize DP, \frac{m\pi}{p\pi + 2\left(1-\pi\right)}\right)
\] 
and therefore the p-value for a two-tailed test is
\[
p = \sum_{i\in I} Bin\left(i \;\large\mid\;\normalsize DP, \frac{m\pi}{p\pi + 2\left(1-\pi\right)}\right)
\]

where $I = \left\{i:P(X=i)\leq P(X=NV)\right\}$.

> The Beta-Binomial case simply follows $P(X_{BB} \leq x) = \sum_{k = 0}^x \textit{Beta-Bin}(x \mid c, \pi/2, \rho)$ for some over-dipsersion $\rho$. A Beta-Binomial distribution can be used to take into account the over-dispersion of the sequencing assay. 

## Example 

In the following example, the classification task is run using both the Binomial and Beta-Binomial models - do appreciate the difference across the width of the acceptance regions of each class. 

The output contains the input plus a `classifier` list that contains, for each model
used for the classification, the parameters of
the fit and the outcome of the classifier for all the possible classes,
along with the corresponding p-values.

```{r}
data = run_classifier(
  x = example_data,
  alpha_level = 0.1,
  model = "Binomial"
)

data = run_classifier(
  x = data,
  alpha_level = 0.1,
  model = "Beta-Binomial",
  rho = 0.01
)

print(data)
```
The results of classification for mutations affecting a specific gene in the panel 
can be visualized using the internal plotting functions. First plots must be
generated.
```{r}
data = plot_test(data)
```
The following example shows
results for a specific gene, "TP53".
```{r, fig.height=4, fig.width=8, dpi=300}
plot_gene(data, model = "Binomial", gene_name = "TP53")
plot_gene(data, model = "Beta-Binomial", gene_name = "TP53")
```
The test shows that at fixed $\alpha$-level, including over-dispersion results in 
a larger confidence interval on $NV$ so that it is more unlikely that karyotype-multiplcity 
hypotheses are rejected.