---
title: "Classify mutations and plot results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(TAPACLOTH)
```

`TAPACLOTH` can be used to classify mutations from targeted panel sequencing data.
First, make sure that input data is in the requested form. It should be a tibble
with `sample`(sample name), `gene`(name of the target gene), `nv` (number of reads
with variant), `dp` (coverage), `vaf` (variant allele frequency) and `purity`
(sample purity) columns. For example:

```{r}
data = dplyr::tibble(sample = "test",
           gene = c("test gene 1", "test gene 2", "test gene 3", "target gene"),
           nv = c(10, 50, 90, 120),
           dp = c(100, 100, 100, 200),
           VAF = c(0.1, 0.5, 0.9, 0.6),
           purity = 1
)
```
Then, you can run the classification on a sample using the Beta-Binomial model:
```{r}
classified_data = run_classifier(
  data = data,
  alpha_level = 1e-3,
  model = "Beta-Binomial",
  rho = 0.01
)
```
The object `classified_data` now contains a `fit` object that is basically the same
input table with the additional column `class_binom`
```{r}
classified_data$fit
```
plus the input parameters of the used test. 

Now you can plot the classification results for the gene you are interested in:
```{r}
plot_tapacloth(fit = classified_data, target_gene = "target gene", sample_name = "test")
```
The classification can also be run per gene using a terzile test. Suppose we have
calls for the same target gene across different samples:
```{r}
data = dplyr::tibble(sample = paste0("test ", seq(1:20)),
           gene = "target gene",
           nv = seq(10,100,4.6) %>% round(0),
           dp = 100,
           VAF = (seq(10,100,4.6) %>% round(0))/100,
           purity = 1
)
```

```{r}
classified_data = run_classifier(
  data = data,
  model = "terzile"
)

classified_data$fit
```
Now you can plot the classification results for the gene and the sample you are interested in:

```{r}
plot_tapacloth(fit = classified_data, target_gene = "target gene", sample_name = "test 1")
```
