---
title: "Analysis of metastatic propensity and tropism of the MSK-MetTropism Cohort"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{metastasis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dplyr)
library(INCOMMON)
```

In this example we perform an analysis of metastatic propensity and tropism using INCOMMON 
classification of 2484 samples from breast cancer (BRCA) patients of the MSK-MetTropsim cohort.

```{r}
data = data_MSK %>% filter(tumor_type == 'BRCA')
print(data)
print(paste0('N samples = ', data$sample %>% unique() %>% length()))
```
## Classification of 2484 MSK-MetTropism BRCA samples

First, we classify all the mutations in these samples, exploiting prior knowledge
from PCAWG. In order to retrieve groups with a large enough number of patients to 
achieve statistical significance, we use no cut-off on entropy.

```{r, message=FALSE}
classified_data = lapply((data$sample %>% unique()), function(s){
  
  what = data %>% filter(sample == s)

  x = init(
    mutations = what,
    sample = s,
    tumor_type = unique(what$tumor_type),
    purity = unique(what$purity),
    gene_roles = cancer_gene_census
  )
  
  x = classify(
    x = x,
    priors = pcawg_priors,
    entropy_cutoff = 1,
    rho = 0.01
  )
  
  output = x
  return(output)
})
```

## Logistic regression test of metastatic propensity 

For the analysis of metastatic propensity, we select only primary tumour samples 
and separate them into samples from metastatic (`metastatic` = 1) and non metastatic 
(`metastatic` = 0) patients. We also keep only genes with Tier-1 mutations in at least 100 samples.

```{r}
what = classification_cohort(classified_data) %>% 
  INCOMMON:::reduce_classes() %>% 
  dplyr::filter(class != 'Tier-2') %>% 
  dplyr::filter(SAMPLE_TYPE == 'Primary') %>% 
  dplyr::mutate(metastatic = 1*(MET_COUNT > 0)) %>% 
  dplyr::select(sample, tumor_type, gene, class, metastatic)

what = what %>% 
  dplyr::group_by(tumor_type, gene) %>% 
  dplyr::reframe(sample, N = length(unique(sample)), class, metastatic) %>% 
  dplyr::filter(N >= 100)

print(what)
```
For each gene, we perform logistic regression using the `stast::glm` function for generalized linear model
fitting. We then adjust p-values using the Benjamini-Hochberg correction for multiple hypotheses testing.

```{r}
out = lapply(what$gene %>% unique(), function(g){
  
  what = what %>%
    dplyr::filter(gene == g) 
  
  what = what %>%
    dplyr::mutate(class = factor(class)) %>%
    dplyr::mutate(class = stats::relevel(class, ref = grep('without', unique(what$class), value = T)))
  
  model = stats::glm(data = what, metastatic ~ class, family = binomial(link = 'logit'))
  
  fit = cbind(summary(model)$coefficients, confint(model)) %>% as_tibble()
  fit$var = rownames(confint(model)) %>% gsub('class', '', .)
  
  dplyr::tibble(
    gene = g,
    N = unique(what$N),
    class = fit[2,]$var,
    OR = fit[2,]$Estimate %>% exp(),
    low = fit[2,]$`2.5 %` %>% exp(),
    up = fit[2,]$`97.5 %` %>% exp(),
    p.value = fit[2,]$`Pr(>|z|)`
  )
}) %>% do.call(rbind, .)

out$p.adjust = stats::p.adjust(out$p.value, method = 'BH')

print(out)
```

Finally, we visualize the fitted odds ratios (OR) in a volcano plot.

```{r}
 plot_volcano = out %>% 
  ggplot2::ggplot(ggplot2::aes(x = log2(OR), y = -log10(p.value)))+
  ggplot2::geom_hline(data = out %>% filter(p.value <= 0.05),
             yintercept = -log10(0.05),
             linetype = 'longdash',
             color = 'indianred3')+
  ggplot2::geom_vline(xintercept = log2(.5),
             linetype = 'longdash',
             color = 'indianred3')+
  ggplot2::geom_vline(xintercept = log2(2),
             linetype = 'longdash',
             color = 'indianred3')+
  ggplot2::geom_point(ggplot2::aes(size = N, color = class))+
  scale_color_INCOMMON_high_level_class(aes = 'color')+
  ggplot2::xlab('Odds Ratio (log2)')+
  ggplot2::ylab('Adjusted p-value (log10)')+
  CNAqc:::my_ggplot_theme(cex = .8)+
  ggrepel::geom_label_repel(data = out %>% filter(p.value<=0.05),
                            ggplot2::aes(label = gene),
                            size = 2.5,
                            min.segment.length = 0.3,
                            force = 150,
                            ylim = c(1.5, NA), show.legend = F)

plot_volcano
```

## Logistic regression test of metastatic tropism

For the analysis of metastatic tropism, we select only tumour samples from metastatic patients
whose metastatic site is specified. 
We also keep only genes with Tier-1 mutations in at least 100 samples, and restrict the analysis
to sites which at least 100 samples metastasize to.

```{r}
what = classification_cohort(classified_data) %>% 
  reduce_classes() %>% 
  dplyr::filter(class != 'Tier-2') %>% 
  dplyr::filter(!is.na(METASTATIC_SITE), METASTATIC_SITE != 'Unspecified') %>%
  dplyr::select(sample, tumor_type, gene, class, METASTATIC_SITE)

genes = what %>% 
  dplyr::group_by(tumor_type, gene) %>% 
  dplyr::reframe(N = length(unique(sample))) %>% 
  dplyr::filter(N >= 100) %>% dplyr::pull(gene) %>% unique()

met_sites = what %>% 
  dplyr::group_by(tumor_type, METASTATIC_SITE) %>% 
  dplyr::reframe(N = length(unique(sample))) %>% 
  dplyr::filter(N >= 100) %>% dplyr::pull(METASTATIC_SITE) %>% unique()

what = what %>% 
  dplyr::filter(gene %in% genes, METASTATIC_SITE %in% met_sites)

print(what)
```

For each gene, we perform logistic regression using the `stast::glm` function for generalized linear model
fitting. We then adjust p-values using the Benjamini-Hochberg correction for multiple hypotheses testing.

```{r}
out = lapply(what$gene %>% unique(), function(g) {
  lapply(what$METASTATIC_SITE %>% unique(), function(m) {
    
    what = what %>%
      dplyr::filter(gene == g) %>%
      dplyr::mutate(tropic = 1 * (METASTATIC_SITE == m))
    
    what = what %>%
      dplyr::mutate(class = factor(class)) %>%
      dplyr::mutate(class = stats::relevel(class, ref = grep('without', unique(what$class), value = T)))
    
    model = stats::glm(data = what,
                       tropic ~ class,
                       family = binomial(link = 'logit'))
    
    
    fit = cbind(summary(model)$coefficients, confint(model)) %>% as_tibble()
    fit$var = rownames(confint(model)) %>% gsub('class', '', .)
    
    dplyr::tibble(
      gene = g,
      METASTATIC_SITE = m,
      N = unique(what$N),
      class = fit[2, ]$var,
      OR = fit[2, ]$Estimate %>% exp(),
      low = fit[2, ]$`2.5 %` %>% exp(),
      up = fit[2, ]$`97.5 %` %>% exp(),
      p.value = fit[2, ]$`Pr(>|z|)`
    )
  }) %>% do.call(rbind, .)
}) %>% do.call(rbind, .)


out$p.adjust = stats::p.adjust(out$p.value, method = 'BH')

print(out)
```
Finally, we visualize the fitted odds ratios (OR) in a volcano plot.

```{r}
plot = out %>%
  dplyr::rowwise() %>%
  dplyr::mutate(
    prevalent = dplyr::case_when(
      p.value <= .05 & OR < 1 ~ 'without LOH',
      p.value <= .05 & OR >= 1 & class == 'with LOH' ~ 'with LOH',
      p.value <= .05 & OR >= 1 & class == 'with AMP' ~  'with AMP',
      TRUE ~ 'ns'
    )
  ) %>%
  ggplot2::ggplot(ggplot2::aes(x = log2(OR), y = gene)) +
  ggplot2::geom_vline(xintercept = 0,
                      linetype = 'longdash',
                      color = 'indianred') +
  
  ggplot2::geom_errorbar(ggplot2::aes(
    xmin = log2(low),
    xmax = log2(up),
    color = prevalent
  ), width = .2) +
  ggplot2::geom_point(ggplot2::aes(fill = prevalent, color = prevalent), shape = 21) +
  scale_color_INCOMMON_high_level_class(aes = 'fill') +
  scale_color_INCOMMON_high_level_class(aes = 'color') +
  ggplot2::xlab('Odds Ratio (log2)') +
  ggplot2::ylab('') +
  ggplot2::facet_wrap(~ METASTATIC_SITE) +
  CNAqc:::my_ggplot_theme(cex = 1)+
  ggplot2::guides(color  = ggplot2::guide_legend(title = 'Class'),
                  fill  = ggplot2::guide_legend(title = 'Class'))

plot
```
Only TP53 and CDH1 present a significant association between the mutant with LOH state and 
a tropism towards liver and bone metastases, respectively.
