---
title: "Survival analysis of the MSK-MetTropism Cohort"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{classify_msk}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(INCOMMON)
library(dplyr)
```

In this example we perform survival analysis using INCOMMON classification of 1742
samples from pancreatic adenocarcinoma (PAAD) patients of the MSK-MetTropsim cohort.

```{r}
data = data_MSK %>% filter(tumor_type == 'PAAD')
print(data)
print(paste0('N samples = ', data$sample %>% unique() %>% length()))
```

First, we classify all the mutations in these samples, exploiting prior knowledge
from PCAWG. In order to retrieve groups with a large enough number of patients to 
achieve statistical significance, we use no cut-off on entropy.

```{r, message=FALSE}
classified_data = lapply((data$sample %>% unique()), function(s){
  
  what = data %>% filter(sample == s)

  x = init(
    mutations = what,
    sample = s,
    tumor_type = unique(what$tumor_type),
    purity = unique(what$purity),
    gene_roles = cancer_gene_census
  )
  
  x = classify(
    x = x,
    priors = pcawg_priors,
    entropy_cutoff = 1,
    rho = 0.01
  )
  
  output = x
  return(output)
})
```
Here, we want to estimate the prognostic power of KRAS mutations with amplification of the mutant allele.  
We first look at the distribution of INCOMMON classes across PAAD samples:

```{r}
plot_class_fraction(x = classified_data, tumor_type = 'PAAD', gene = 'KRAS')
```

Nearly 30% of KRAS mutations are associated with amplification of the mutant allele, through CNLOH.
Next we stratify PAAD patients as KRAS WT, Mutant KRAS without amplification and Mutant KRAS with amplification, and
we fit survival data (overall survival status versus overall survival months) using the Kaplan-Meier estimator.

```{r}
km_fit = kaplan_meier_fit(x = classified_data, tumor_type = 'PAAD', gene = 'KRAS')
km_fit$fit[[1]]
```

The median survival time decreases from 21.5 months for the KRAS WT group to 18.2 months
for Mutant KRAS without amplification and further to 11.7 months for Mutant KRAS with amplification
patients.

In order to estimate the hazard ratio associated with these groups, we fit the same survival 
data, this time using a multivariate Cox proportional hazards regression model. Here, we include
the age of patients at death, sex and tumor mutational burden as model covariates.

```{r}
cox_fit(x = classified_data, tumor_type = 'PAAD', gene = 'KRAS', covariates = c('age', 'sex', 'tmb'))
```

This analysis reveals that, whereas KRAS mutation alone (without amplification) is not 
enough, the presence of amplification significantly increases the hazard ratio (HR = 1.43, p-value = 0.009) 
with respect to the WT group. Moreover, age seems to play a significant role, as patients younger 
than 68 (median age) emerge as being more at risk (HR = 0.87, p-value = 0.036) than the older ones.

Kaplan-Meier estimation and multivariate Cox regression can be run and visualized straightforwardly
using the following function.

```{r}
plot_survival_analysis(x = classified_data, tumor_type = 'PAAD', gene = 'KRAS', cox_covariates = c('age', 'sex', 'tmb'))
```
The plot displays Kaplan-Meier survival curves and risk table, and the forest plot 
for the Cox regression coefficients, where significant covariates are shown in red. 

